---
title: "Apply hierarchical consensus partitioning on the example dataset in Figure 1"
author: "Zuguang Gu (z.gu@dkfz.de)"
date: '`r Sys.Date()`'
output:
  html_document:
    fig_caption: true
---

```{r, echo = FALSE}
library(knitr)
knitr::opts_chunk$set(
    warning = FALSE,
    results = FALSE,
    fig.align = "center")
```

In Figure 1 of the manuscript, we constructed a random matrix that contained groups with large difference
as well as groups with small difference and we demonstrated that standard consensus partitioning procedures
cannot separate all the four groups simultaneously. In this supplementary, we demonstrate that hierarchical consensus
partitioning is able to identify all groups.


First we generate the random matrix the same as in Figure 1.

```{r, fig.cap = "Figure S1.1. Heatmap of the random matrix used in Figure 1 in the manuscript."}
library(ComplexHeatmap)

set.seed(54)
mean_diff1 = rnorm(100)

m1 = do.call(rbind, lapply(1:100, function(i) {
	c(rnorm(50, mean = mean_diff1[i]), rnorm(50, mean = -mean_diff1[i]))
}))

mean_diff2 = rnorm(100)/2
mean_diff2[order(abs(mean_diff1))] = mean_diff2[order(abs(mean_diff2), decreasing = TRUE)]

m2 = do.call(rbind, lapply(1:100, function(i) {
	c(rnorm(10, mean = mean_diff2[i]), rnorm(10, mean = -mean_diff2[i]))
}))

m = cbind(m1, m2)

group = rep(c("A1", "A2", "B1", "B2"), times = c(50, 50, 10, 10))
group_col = structure(c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3"), names = c("A1", "A2", "B1", "B2"))

Heatmap(m, name = "mat", 
	top_annotation = HeatmapAnnotation(Group = group, col = list(Group = group_col)),
	show_row_dend = FALSE, column_title = "Heatmap of the random dataset",
	row_dend_reorder = mean_diff1, column_dend_reorder = as.numeric(factor(group))
)
```

The standard consensus partition procedures cannot identify all four groups. It can only identify
three groups as the best results:

```{r, fig.width = 10, fig.height = 10, fig.cap = "Figure S1.2. Various plots for interpreting consensus partitioning results."}
res = consensus_partition(m, top_value_method = "SD", partition_method = "kmeans",
	top_n = 50, anno = group, anno_col = group_col, scale_rows = FALSE)
collect_plots(res)
```

Then we apply hierarchical consensus partitioning with the function `hierarchical_partition()` on the matrix:

```{r}
library(cola)
rh = hierarchical_partition(m, 
	top_value_method = "SD", partition_method = "kmeans", 
	anno = group, anno_col = group_col,
	top_n = 50, scale_rows = FALSE)
```

We can print the `rh` object:

```{r, results = TRUE}
rh
```

The function `collect_classes()` draws the subgroup hierarchy:

```{r, fig.cap = "Figure S1.3. Hierarchiy of cola HCP."}
collect_classes(rh)
```

As we can see, all the four groups are separated.

## Session info

```{r, results = TRUE}
sessionInfo()
```
