---
title: "Automatically select the 'knee/elbow' of a curve"
author: "Zuguang Gu (z.gu@dkfz.de)"
date: '`r Sys.Date()`'
output:
  html_document:
    fig_caption: true
---

```{r, echo = FALSE}
library(knitr)
knitr::opts_chunk$set(
    warning = FALSE,
    message = FALSE,
    results = FALSE,
    fig.align = "center")
```

We demonstrate the method that automatically selects the "knee/elbow" of a curve proposed in the study [Satoppa et al., 2011](https://ieeexplore.ieee.org/document/5961514).

We first get a vector of scores which are the row means of the matrix from the Golub dataset.

```{r}
library(cola)
library(golubEsets)
data(Golub_Merge)
m = exprs(Golub_Merge)

m[m <= 1] = NA
m = log10(m)
m = adjust_matrix(m)

library(matrixStats)
s = rowSds(m)
```

Values in `s` are sorted increasingly and they are visualized as a
scatterplot. Now the task is to identify the "elbow" of this curve.

```{r, fig.cap = "Figure S2.1. Sorted row means of Golub dataset."}
s = sort(s)
plot(s)
```

[Satoppa et al., 2011](https://ieeexplore.ieee.org/document/5961514) proposed a method that selects the "elbow point" as the one having the
largest distance to the straight line (denoted as $L$) that connects the two boundary points with the minimal and maximal values on _y_-axis. The method is demonstrated by the following code. Note values on both _x_-axis and _y_-axis are scaled into `[0, 1]`.

```{r, fig.cap = "Figure S2.2. Demonstration of the method to look for elbow/knee of the curve."}
y = s
x = seq_along(y)

x = x/max(x)
y = y/max(y)

n = length(x)
a = (y[n] - y[1])/(x[n] - x[1])
b = y[1] - a * x[1]
d = a * x + b - y
x1 = x[which.min(d)]
x2 = x[which.max(d)]
theta = atan(a)

plot(x, y, xlab = "index", ylab = "value", asp = 1)
abline(a = b, b = a)

breaks = seq(1000, 3800, length = 8)/n
y0 = quantile(y, breaks)
y1 = a*breaks + b
segments(breaks, y0, breaks, y1, col = "red")

a2 = tan(theta + pi/2)
b2 = y0- a2*breaks
x2 = (b2 - b)/(a - a2)
y2 = a*x2 + b
segments(breaks, y0, x2, y2, col = "blue")
```

Instead of selecting by the distance to $L$ (the lengths of blue lines), we can simply select according
to the vertical distance to $L$ (the lengths of red lines). The two selection rules are actually identical.

We implement this method with the function `knee_finder2()` in **cola**. The function generates two plots: the original curve
and a curve of the vertical distance to $L$.


```{r, fig.width = 10, fig.height = 5, results = TRUE, fig.cap = "Figure S2.3. Output plots from knee_finder2()."}
knee_finder2(s, plot = TRUE)
```

`knee_finder()` returns a vector of two. The second value corresponds to the elbow of the curve
and the first value corresponds to the knee of the curve, as shown in the following plot.


```{r, fig.width = 10, fig.height = 5, results = TRUE, fig.cap = "Figure S2.4. When both knee and elbow exist in the curve."}
s = rnorm(1000)
knee_finder2(s, plot = TRUE)
```

## Session info

```{r, results = TRUE}
sessionInfo()
```