---
title: "Analysis of the TCGA GBM microarray dataset"
author: "Zuguang Gu (z.gu@dkfz.de)"
date: '`r Sys.Date()`'
output:
  html_document:
    fig_caption: true
---

```{r, echo = FALSE}
library(knitr)
knitr::opts_chunk$set(
    warning = FALSE,
    message = FALSE,
    results = "hide",
    fig.align = "center")
```

```{r, echo = FALSE}
library(cola)
```

We read the matrix and perform normalization.


```{r, eval = FALSE}
library(cola)

m = read.table("https://jokergoo.github.io/cola_examples/TCGA_GBM/unifiedScaled.txt", 
    header = TRUE, row.names = 1, check.names = FALSE)
m = as.matrix(m)

subtype = read.table("https://jokergoo.github.io/cola_examples/TCGA_GBM/TCGA_unified_CORE_ClaNC840.txt", 
    sep = "\t", header = TRUE, check.names = FALSE, stringsAsFactors = FALSE)
subtype = structure(unlist(subtype[1, -(1:2)]), names = colnames(subtype)[-(1:2)])
subtype_col = structure(seq_len(4), names = unique(subtype))

m = m[, names(subtype)]
m = adjust_matrix(m)

library(preprocessCore)
cn = colnames(m)
rn = rownames(m)
m = normalize.quantiles(m)
colnames(m) = cn
rownames(m) = rn
```

First we apply standard consensus partitioning analysis with "ATC" as the top-value method
and "skmeans" as partitioning method.

```{r, eval = FALSE}
set.seed(123)
res = consensus_partition(m, top_value_method = "ATC", partition_method = "skmeans",
    cores = 4, anno = subtype, anno_col = subtype_col)
```

```{r, echo = FALSE}
res = readRDS("TCGA_GBM_microarray_CP.rds")
```

In the following plot, **cola** suggests 5 as the best number of subgroups, but
we select 4 as the best k because it gives more stable classification.

```{r, results = "markup", message = TRUE}
suggest_best_k(res)
```

```{r, fig.width = 8, fig.height = 8, fig.cap = "Figure S6.1. Select the best number of groups."}
select_partition_number(res)
```

The signature heatmap with 4 subgroups.

```{r, fig.width = 8, fig.height = 8, fig.cap = "Figure S6.2. Signature heatmap of CP classification with 4 subgroups."}
get_signatures(res, k = 4)
```

Next we apply hierarchical consensus partitioning (HCP) on the same matrix:

```{r, eval = FALSE}
set.seed(123)
rh = hierarchical_partition(m, cores = 4, anno = subtype, anno_col = subtype_col)
```

```{r, echo = FALSE}
rh = readRDS("TCGA_GBM_microarray_HCP.rds")
```

The subgroup hierarchy:

```{r, fig.height = 8, fig.cap = "Figure S6.3. Subgroup hierarchy under HCP."}
collect_classes(rh)
```

And the signature heatmap under HCP classification:

```{r, fig.width = 8, fig.height = 8, fig.cap = "Figure S6.4. Signature heatmap under HCP classification."}
get_signatures(rh)
```

The statistics on each node:

```{r, results = "markup"}
df = node_info(rh)
df
```

And the statistics on non-leaf nodes:

```{r, results = "markup"}
df[!df$is_leaf, ]
```

## Session Info

```{r, results = "markup"}
sessionInfo()
```